Finiteness in Cubical Type Theory

### 2020-05-29 paper.tex
split into sections

%   We formalise our work in Cubical Agda, where we implement a library for proof
%   search (including combinators for level-polymorphic fully generic currying).
%   Through this library we demonstrate a number of uses for the computational
%   content of the univalence axiom, including searching for and synthesising
%   functions.
% \end{abstract}

\section{Introduction}
\subsection{Foreword}
Proofs in constructive mathematics are often more substantial than their
classical counterparts.
``Substantial'' here doesn't mean the literal length of the proofs (although the
double meaning is not lost on anyone who has had to struggle with a modern
constructive proof assistant), but rather refers to their \emph{content}: a
constructive proof contains more \emph{stuff} than a classical one.

If we were to prove that a set was finite in classical mathematics, we might
show that it is in bijection with another finite set, say a finite prefix of the
natural numbers.
In semi-formal notation, to say that a set \(A\) is finite we would show the
following:
\begin{equation} \label{example-finite}
  \mathbf{Finite}(A) \coloneqq \exists (n : \mathbb{N}) . \; A \Leftrightarrow \left\{ m \mid m < n\right\}
\end{equation}
Where \(A~\Leftrightarrow~B\) means ``a bijection between \(A\) and \(B\)''.

Classically, a proof of equation~\ref{example-finite} is nothing more than
\emph{evidence} for the proposition of equation~\ref{example-finite}.
A constructive proof of \ref{example-finite}, however, is a mathematical object
in its own right, containing:
\begin{enumerate}
  \item A number \(n\) representing the cardinality of \(A\).
  \item A function from \(A\) to \(\left\{ m \mid m < n \right\}\).
  \item A function from \(\left\{ m \mid m < n \right\}\) to \(A\).
\end{enumerate}
In other words, it makes sense (in constructive mathematics) to talk about a
function of type \(\mathbf{Finite}(A)~\rightarrow~\mathbb{N}\), which takes a
\emph{proof} of the finiteness of \(A\) and returns \(A\)'s cardinality.

Such functions are perhaps even more concrete than this description suggests:
many constructive proofs are in fact computer programs in disguise (and
vice-versa \cite{wadlerPropositionsTypes2015}).
All ``proofs'' in this paper correspond to honest-to-goodness proofs in a
programming language (Cubical Agda \cite{vezzosiCubicalAgdaDependently2019} in
our case) which can be run on a computer to produce real output.

We have now hopefully established, albeit informally, what it means for
constructive proofs to be ``substantial''.
Clearly this extra information carried by proofs can add richness to their
study: we will see, for instance, no fewer than five finiteness predicates in
this paper which would collapse to one in a classical setting.
The extra information also presents a challenge, however: occasionally we will
see that we have revealed \emph{too much} information while proving some
property.

Consider the category of finite sets.
We will explore this category in much more detail later, but for now we just
need to know that the objects of this category are---shockingly---finite sets.
Constructively, the objects of this category can be thought of as \emph{pairs},
where the first element is a set (like \(\left\{ \text{True}, \text{False}
\right\}\), the two-element set of the booleans) and the second is a proof that
that set is finite.
If we take those proofs to be the kind given in equation~\ref{example-finite},
we should see that they contain a bijection from the booleans to the set
\(\left\{ 0, 1 \right\}\): however \emph{two} such bijections exist!
\begin{figure}[h]
  \centering
  \begin{subfigure}[b]{.45\textwidth}
    \centering
    \begin{tikzcd}
      \text{True}  \ar[rr, leftrightarrow] & & 0 \\
      \text{False} \ar[rr, leftrightarrow] & & 1
    \end{tikzcd}
  \end{subfigure}
  \begin{subfigure}[b]{.45\textwidth}
    \centering
    \begin{tikzcd}
      \text{True}  \ar[rrd, leftrightarrow, start anchor=south east, end anchor=north west] & & 0 \\
      \text{False} \ar[rru, leftrightarrow, start anchor=north east, end anchor=south west, crossing over] & & 1
    \end{tikzcd}
  \end{subfigure}
  \caption{The Two Bijections Between the Booleans and \(\left\{ 0, 1 \right\}\)}
\end{figure}
This means that where we wanted one object to correspond to the finite set
\(\left\{ \text{True}, \text{False} \right\}\), we instead have two.
In fact, for any finite set of cardinality \(n\) we will have \(n!\) finiteness
proofs, and therefore \(n!\) objects.
We have constructed, in other words, entirely the wrong category.

Being able to hide the extra information in a constructive proof when we need to
is clearly important.
HoTT (and CuTT by extension) facilitate this hiding: how exactly it is
accomplished will be explored in the subsequent sections.

Finally, for all of its benefits, constructive mathematics (especially
dependently-typed constructive mathematics on a computer) still suffers from
many painful drawbacks.
As already hinted at, proofs in a constructive setting can tend to be verbose,
especially when the arbiter of correctness is as strict as a computer, which
neither understands nor trusts the time-honoured technique of ``the other cases
are obviously true''.
While the goal of using a computer to automate the boring parts of a proof still
seems attainable, too often getting a computer involved in a proof means adding
thousands of lines of tedium to satisfy its lack of intuition.

Simple proof search over a finite domain is a simple, if effective, way to
relieve some of the tedium.
Many libraries in Agda, Coq, and other languages make use of this fact.
While the idea for finite proof search is simple, we found that implementing one
revealed a number of interesting theoretical considerations:
\begin{description}
  \item[How should we construct search spaces?]
    We found that a \(Pi\)-pretopos provided the appropriate ``generator
    language'' for constructing search spaces for finite types.
  \item[How do we deal with functions?]
    We of course would like to be able to search over functions: \todo{?}
  \item[How do we talk about ``searchability''?]
    \todo{Omniscience}
\end{description}
Finiteness is a prime example of a phenomenon which happens quite frequently in
dependently-typed constructive mathematics: through building a practical program
we unearth theoretically interesting results.
\todo{rephrase}
\subsection{Contributions}
We are interested in constructive notions of finiteness, their classification as
a topos, their relation to countability, and their use as a proof search library
in Cubical Agda.
\subsubsection{Finiteness Predicates}
Our focus in this work is decidable Kuratowski finiteness.
We will characterise it, its strength, and its uses, and in particular we will
show that it forms a \(\Pi\)-pretopos.
Our proofs follow the structure of \cite[Chapters 9, 10]{hottbook} and
\cite{rijkeSetsHomotopyType2015}.
We will explore this concept in more detail later, but crucially we will need to
provide proofs of closure under \(\times\), \(\rightarrow\), etc.

In order to prove these closure properties, we will define and use some
intermediate finiteness predicates, classified in
figure~\ref{finite-classification}.
These predicates are also interesting in their own right, though!

The arrows in figure~\ref{finite-classification} are implication functions
between each predicate.
In section~\ref{relations} we will provide a function which inhabits each arrow.
Each unlabelled arrow is an unconditional implication: every manifest Bishop
finite set is cardinal finite (lemma~\ref{manifest-bishop-to-cardinal}), for
instance.
The labelled arrows are strengthening proofs: every manifest enumerable set
\emph{with decidable equality} is split enumerable
(lemma~\ref{manifest-enum-to-split-enum}).
Our most significant result here is the proof that a cardinal finite set with a
decidable total order is manifestly Bishop finite.

\input{figures/finite-classification}
\subsubsection{Countability Predicates}
After the finite predicates, we will briefly look at the infinite countable
types, and classify them in a parallel way to the finite predicates
(section~\ref{infinite}).
\subsubsection{Search}
All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in
section~\ref{search}, where we implement a library for proof search, based on
omniscience and exhaustibility.
This library relies directly on the computational content of univalence, which
allows us to (for instance) have \(\Pi\) types in the domain of the search
space.
The language of a \(\Pi\)-pretopos turns out to be exactly the right language
for constructing ``generator''-like expressions.

The full code of the formalisation is available at
\url{https://github.com/oisdk/finiteness-in-cubical-type-theory}.
\subsection{Notation and Background}
\subsubsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into \(\mathbf{Type}\).
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
    The ``negation'' of a type, written \(\neg A\), means \(A \rightarrow
    \bot\).
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] We define disjoint union as an inductive type.
    \begin{equation}
      \begin{alignedat}{3}
        A \uplus B \coloneqq & \;
        \text{inl} &: A \rightarrow A \uplus B ; \\
        | & \;  \text{inr} &: B \rightarrow A \uplus B ;
      \end{alignedat}
    \end{equation}
    It is also expressible with only \(\Sigma\):
    \(A \uplus B \simeq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then
    } A \text{ else } B \).
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage: we will only use one
    or the other as a suggestion of how we constructed it or how it is to be
    used.
\end{description}
\subsubsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, giving it computational content.
\begin{definition}[Path Types] \label{path-types}
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{definition}

\footnotetext{
  Actually, CuTT does have an identity type with similar semantics to the
  identity type in MLTT.
  We do not use this type anywhere in our work, however, so we will not consider
  it here.
}
\begin{definition}[Homotopy Levels] \label{homotopy-types}
  Types in HoTT and CuTT are not necessarily sets, as they are in MLTT.
  Some have higher homotopies (paths which aren't unique).
  We actually have a hierarchy of complexity of structure of path spaces in
  types, starting with the contractions \cite[definition 3.11.1]{hottbook}, then
  the mere propositions \cite[definition 3.3.1]{hottbook}, and the sets
  \cite[definition 3.1.1]{hottbook}.
  \begin{alignat}{2}
    &\text{isContr}(A)    &&\coloneqq \Sigma(x : A) , \Pi(y : A) , (x \equiv y) \\
    &\text{isProp}(A)     &&\coloneqq \Pi(x, y : A) , (x \equiv y) \\
    &\text{isSet}(A)      &&\coloneqq \Pi(x, y : A) , \text{isProp}(x \equiv y)
  \end{alignat}
\end{definition}
\begin{definition}[Fibres] \label{fibres}
  A fibre \cite[definition 4.2.4]{hottbook} is defined over some function \(f :
  A \rightarrow B\).
  \begin{equation}
    \text{fib}_f(y) \coloneqq \Sigma(x : A) , (f (x) \equiv y)
  \end{equation}
\end{definition}
\begin{definition}[Equivalences] \label{equivalences}
  We will take contractible maps \cite[definition 4.4.1]{hottbook} as our
  ``default'' definition of equivalences.
  \begin{alignat}{2}
    &\text{isEquiv}(f) &&\coloneqq \Pi(y : B) , \text{isContr}(\text{fib}_f(y)) \label{is-equiv-def} \\
    &A \simeq B        &&\coloneqq \Sigma(f : A \rightarrow B) , \text{isEquiv}(f)
  \end{alignat}
\end{definition}
\begin{definition}[Decidable Types]
  \begin{equation}
    \mathbf{Dec}(A) \coloneqq A \uplus \neg A
  \end{equation}
\end{definition}
\begin{definition}[Discrete Types]
  A discrete type is one with decidable equality.
  \begin{equation}
    \text{Discrete}(A) \coloneqq \Pi(x, y : A) , \mathbf{Dec}(x \equiv y)
  \end{equation}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998} any discrete
  type is a set.
\end{definition}
\begin{definition}[Higher Inductive Types] \label{HITs}
  Normal inductive types have \emph{point} constructors: constructors which
  construct values of the type.
  Higher Inductive Types (HITs) also have \emph{path} constructors: ways to
  construct paths in the type.
\end{definition}
\begin{definition}[Propositional Truncation] \label{prop-trunc}
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\) \cite[3.7]{hottbook}.
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

  It is defined as a Higher Inductive Type:
  \begin{equation} {
    \begin{alignedat}{3}
      \lVert A \rVert \coloneqq & \; \lvert \wc \rvert &:& \; A \rightarrow \lVert A \rVert ; \\
                              | & \; \text{squash}     &:& \; \Pi {(x, y : \lVert A \rVert)} , x \equiv y  ; 
    \end{alignedat} }
  \end{equation}
  We will use two eliminators from \(\lVert A \rVert\) in this paper.
  \begin{enumerate}
  \item \label{elim-prop-prop} For any function \(A \rightarrow B\), where
    \(\text{isProp}(B)\), we have a function \(\lVert A \rVert \rightarrow B\).
  \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
    function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
    choice of \(A\):
    \[\Pi {(x , y : A)} , f(x) \equiv f(y) \]
    Formally speaking, \(f\) needs to be ``coherently constant''
    \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
    \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{definition}
\section{Finiteness Predicates} \label{finiteness-predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
The reason we explore predicates other than our focus (decidable Kuratowski
finiteness) is that we can often prove things like closure much more readily on
the simpler predicates.
The relations (which we will prove in the next section) then allow us to
transfer those proofs.
\subsection{Split Enumerability}
\begin{definition}[Split Enumerable Set] \label{split-enum-def}
  \begin{equation} \label{split-enum-def-eqn}
    \mathcal{E}!(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{definition}

We used some extra types in the above definition, which we will define here:
\begin{definition}[Containers] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S , P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{equation} \label{container-interp}
    \llbracket S , P \rrbracket(A) \coloneqq \Sigma {(s : S)} , \left( P(s) \rightarrow A \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation} \label{container-membership}
    x \in \mathit{xs} \coloneqq \text{fib}_{\text{snd}(\mathit{xs})}(x)
  \end{equation}
\end{definition}
\begin{definition}[\(\mathbf{List}\)] \label{List}
  \begin{equation}
    \mathbf{List} \coloneqq \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{definition}
\begin{definition}[\(\mathbf{Fin}\)] \label{Fin}
  \(\mathbf{Fin}(n)\) is the type of natural numbers smaller than \(n\). We
  define it the standard way:
  \begin{equation}
    \begin{alignedat}{3}
      & \mathbf{Fin}(0)     && \coloneqq \bot ; \\
      & \mathbf{Fin}(n + 1) && \coloneqq \top \uplus \mathbf{Fin}(n) ;
    \end{alignedat}
  \end{equation} 
\end{definition}

We tend to prefer list-based definitions of finiteness, rather than ones based
on bijections or surjections.
This is purely a matter of perspective, however: the definition above is
precisely equivalent to a split surjection from a finite prefix of the natural
numbers.
\begin{definition}[Surjections] \label{surjections}
  We define both surjections and \emph{split} surjections here \cite[definition
  4.6.1]{hottbook}.
  \begin{alignat}{3}
    &\text{surj}(f)             &&\coloneqq \Pi(y : B) , \lVert \text{fib}_f(y) \rVert \\
    &A \twoheadrightarrow B     &&\coloneqq \Sigma (f : A \rightarrow B) , \text{surj}(f) \label{surj-arrow-eqn} \\
    &\text{sp-surj}(f)          &&\coloneqq \Pi(y : B) , \text{fib}_f(y) \label{sp-surj-eqn} \\
    &A \twoheadrightarrow! \; B &&\coloneqq \Sigma (f : A \rightarrow B) , \text{sp-surj}(f) \label{sp-surj-arrow-eqn}
  \end{alignat}
\end{definition}
\begin{lemma} \label{split-enum-is-split-surj}
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma (n : \mathbb{N}) , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{E}!(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \in \mathit{xs}
    && \text{def.~\ref{split-enum-def} }(\mathcal{E}!)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{fib}_{\text{snd}(\mathit{xs})}(x)
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{sp-surj-eqn} (sp-surj)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{sp-surj}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{sp-surj}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \twoheadrightarrow ! \; A )
    && \text{eqn.~\ref{sp-surj-arrow-eqn} } (\twoheadrightarrow!)
  \end{align*}
\end{proof}

In our formalisation, the proof is a single line: most of the steps above are
simple expansion of definitions.
The only step which isn't definitional equality is the reassociation of
\(\Sigma\).

Split enumerability implies decidable equality on the underlying type.
To prove this, we will make use of the following lemma, proven in the
formalisation:
\begin{lemma} \label{discrete-surj}
  \begin{equation}
    \frac{
        A \twoheadrightarrow! \; B \; \; \; \text{Discrete}(A)
      }{
       \text{Discrete}(B) 
      }
  \end{equation}
\end{lemma}
\begin{lemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{lemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\mathbf{Fin}(n)\) for some \(n\).
  Also, we know that \(\mathbf{Fin}(n)\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
\end{proof}
\subsection{Manifest Bishop Finiteness}
\begin{definition}[Manifest Bishop Finiteness]
  \begin{equation} \label{bish-def}
    \mathcal{B}(A) \coloneqq \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{definition}
The only difference between manifest Bishop finiteness and split enumerability
is the membership term: here we require unique membership (\(\inunique\)),
rather than simple membership (\(\in\)).
\begin{definition}[Unique Membership] \label{uniq-memb-def}
  \begin{equation}
    x \inunique \mathit{xs} \coloneqq \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{definition}

We use the word ``manifest'' here to distinguish from another common
interpretation of Bishop finiteness, which we have called cardinal finiteness in
this paper.
The ``manifest'' refers to the fact that we have a concrete, non-truncated list
of the elements in the proof.

Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
This proof is effectively the same as that of
lemma~\ref{split-enum-is-split-surj}.
\subsection{Cardinal Finiteness}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

\footnotetext{
  We actually do get a category (a groupoid, even) from manifest Bishop
  finiteness \cite{yorgeyCombinatorialSpeciesLabelled2014}: it's the groupoid of
  finite sets equipped with a linear order, whose morphisms are order-preserving
  bijections.
  We do not explore this particular construction in any detail.
}

To remedy the problem, we will use propositional truncation
(def.~\ref{prop-trunc}).
\begin{definition}[Cardinal Finiteness]
  \begin{equation}
    \mathcal{C}(A) \coloneqq \lVert \mathcal{B}(A) \rVert \simeq \lVert \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \simeq A) \rVert
  \end{equation}
\end{definition}
At first glance, it might seem that we lose any useful properties we could
derive from \(\mathcal{B}\).
Luckily, this is not the case: by eliminator \ref{elim-prop-coh} of
def.~\ref{prop-trunc}, we need only show that the output is uniquely determined.

The following two lemmas are proven in
\cite{yorgeyCombinatorialSpeciesLabelled2014} (Proposition 2.4.9 and 2.4.10,
respectively), in much the same way as we have done here.
Our contribution for this section is simply the formalisation.
\begin{lemma}
  Given a cardinally finite type, we can derive the type's cardinality, as well
  as a propositionally truncated proof of equivalence with \(\textbf{Fin}\)s of
  the same cardinality.
  \begin{equation}
    \mathcal{C}(A) \rightarrow \Sigma {(n : \mathbb{N})} , \lVert \textbf{Fin}(n) \simeq A \rVert
  \end{equation}
\end{lemma}
\begin{lemma} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{lemma}
\subsection{Manifest Enumerability}
\begin{definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) \coloneqq \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{definition}
As with manifest Bishop finiteness, the only difference with this type and split
enumerability is the membership proof: here we have propositionally truncated
it.
This has two effects.
First, it means that this proof represents a true surjection (rather than a
split surjection) from \(\mathbf{Fin}\).
\begin{lemma}
  \begin{equation}
    \mathcal{E}(A) \simeq \Sigma(n : \mathbb{N}) , (\mathbf{Fin}(n) \twoheadrightarrow A)
  \end{equation}
\end{lemma}
The proof for this lemma is similar in structure to
lemma~\ref{split-enum-is-split-surj} and lemma~\ref{bishop-equiv}.

Secondly, it means the predicate does not imply decidable equality.
More significantly, it allows the predicate to be defined over non-set types,
like the circle.
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignedat}{3}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
      | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignedat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\subsection{Kuratowski Finiteness}
The first thing we must define is a representation of subsets.
\begin{definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{equation}
    \begin{alignedat}{3}
      \mathcal{K}(A) \coloneqq&
             \; []                &:& \; \mathcal{K}(A) ; \\
      \vert& \; \wc \dblcolon \wc &:& \; A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A) ; \\
      \vert& \; \text{com}        &:& \; \Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , \\
                                  &&& \; x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs} ; \\
      \vert& \; \text{dup}        &:& \; \Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs} ; \\
      \vert& \; \text{trunc}      &:& \; \Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q ;
    \end{alignedat}
  \end{equation}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

  The next two constructors add extra paths to the type: equations that usage of
  the type must obey.
  These extra paths turn the free monoid into the free \emph{commutative} (com)
  \emph{idempotent} (dup) monoid.

  The final constructor enforces that the type \(\mathcal{K}(A)\) must be a set.
\end{definition}
The Kuratowski finite subset is a free join semilattice (or, equivalently, a
free commutative idempotent monoid).
More prosaically, \(\mathcal{K}\) is the abstract data type for finite sets, as
defined in the Boom hierarchy \cite{boomFurtherThoughtsAbstracto1981,
  bunkenburgBoomHierarchy1994}.
However, rather than just being a specification, \(\mathcal{K}\) is fully usable
as a data type in its own right, thanks to HITs.

Other definitions of \(\mathcal{K}\) exist (such as the one in
\cite{fruminFiniteSetsHomotopy2018}) which make the fact that \(\mathcal{K}\) is
the free join semilattice more obvious.
We have included such a definition in our formalisation, and proven it
equivalent to the one above.

Next, we need a way to say that an entire type is Kuratowski finite.
For that, we will need to define membership of \(\mathcal{K}\).
\begin{definition}[Membership of \(\mathcal{K}\)]
  Membership is defined by pattern-matching on \(\mathcal{K}\).
  The two point constructors are handled like so:
  \begin{equation}
    \begin{alignedat}{2}
      x \in&& \; []                      &\coloneqq \bot ; \\
      x \in&& \; y \dblcolon \mathit{ys} &\coloneqq \lVert x \equiv y \uplus x \in \mathit{ys} \rVert ;
    \end{alignedat}
  \end{equation}
  The \(\text{com}\) and \(\text{dup}\) constructors are handled by proving that
  the truncated form of \(\uplus\) is itself commutative and idempotent.
  The type of propositions is itself a set, satisfying the \(\text{trunc}\)
  constructor.
\end{definition}
Finally, we have enough background to define Kuratowski finiteness.
\begin{definition}[Kuratowski Finiteness]
  \begin{equation}
    \mathcal{K}^{f}(A) = \Sigma {(\mathit{xs} : \mathcal{K}(A))} , \Pi (x : A) , x \in \mathit{xs}
  \end{equation}
\end{definition}

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{lemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{lemma}
\begin{lemma}
  This circle \(S^1\) is Kuratowski finite.
\end{lemma}
\section{Relations Between Each Finiteness Definition} \label{relations}
We will now look at the arrows in figure~\ref{finite-classification}.
\subsection{Split Enumerability and Manifest Bishop Finiteness}
While manifest Bishop finiteness might seem stronger than split enumerability,
it turns out this is not the case.
Both predicates imply the other.
\begin{lemma} \label{manifest-bishop-to-split-enum}
  Any manifest Bishop finite type is split enumerable.
\end{lemma}
\begin{proof}
  To construct a proof of split enumerability from one of manifest Bishop
  finiteness, it suffices to convert a proof of \(x \inunique \mathit{xs}\) to
  one of \(x \in \mathit{xs}\), for all \(x\) and \(\mathit{xs}\).
  Since \(\inunique\) is defined as a contraction of \(\in\), such a conversion
  is simply the \(\text{fst}\) function.
\end{proof}

\begin{lemma} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{lemma}
This proof takes significantly more work.
The ``unique membership'' condition in \(\mathcal{B}\) means that we are not
permitted duplicates in the support list.
The first step in the proof, then, is to filter those duplicates out from the
support list of the \(\mathcal{E}!\) proof: we can do this using the decidable
equality provided by \(\mathcal{E}!\) (lemma~\ref{split-enum-discrete}).
From there, all we need to show is that the membership proof carries over
appropriately.
\subsection{Split Enumerability and Manifest Enumerability}
\begin{lemma} \label{split-enum-to-manifest-enum}
  Any split enumerable type is manifestly enumerable.
\end{lemma}
This lemma is proven by truncating the membership proof in split enumerability.

\begin{lemma} \label{manifest-enum-to-split-enum}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{lemma}

\subsection{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{lemma} \label{manifest-bishop-to-cardinal}
  Any manifest Bishop finite type is cardinal finite.
\end{lemma}
\begin{theorem} \label{cardinal-to-manifest-bishop}
  Any cardinal finite type with a total order is Bishop finite.
\end{theorem}
The proof for this particular theorem is quite involved in the formalisation, so
we only give its sketch here.
First, note that we actually convert to manifest enumerability first: this can
be converted to split enumerability with decidable equality, which is provided
by cardinal finiteness.

Next, we define permutations.
\begin{definition}[List Permutations]
  Two lists are permutations of each other if their membership proofs are all
  equivalent\footnotemark \cite{danielssonBagEquivalenceProofRelevant2012}.
  \begin{equation}
    \mathit{xs} \leftrightsquigarrow \mathit{ys} = \Pi {(x : A)} , x \in \mathit{xs} \simeq x \in \mathit{ys}
  \end{equation}
\end{definition}

\footnotetext{
  The definition in \cite{danielssonBagEquivalenceProofRelevant2012} and our
  formalisation is slightly different: we say permutations are lists with
  \emph{isomorphic} membership proofs.
  The distinction, as it happens, does not affect our work here.
}

Next, we define a sort function which relies on the provided total order.
We further prove the following fact about this sort function:
\begin{equation}
  \Pi(\mathit{xs}, \mathit{ys} : \mathbf{List}(A)) , \mathit{xs} \leftrightsquigarrow \mathit{ys} \rightarrow \text{sort}(\mathit{xs}) \equiv \text{sort}(\mathit{ys})
\end{equation}

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Cardinal Finiteness and Kuratowski Finiteness}
\begin{lemma} \label{cardinal-kuratowski}
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}^f(A) \times \text{Discrete}(A)
  \end{equation}
\end{lemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\mathcal{K}^f(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions.
The actual functions themselves are proven in our formalisation, as well as in
\cite{fruminFiniteSetsHomotopy2018}.
\section{Topos} \label{topos}
In this section we will prove that decidable Kuratowski finite types form a
\(\Pi\)-pretopos.
Along the way we will provide closure proofs for a number of the other
finiteness predicates.
As we saw in theorem~\ref{cardinal-kuratowski}, decidable Kuratowski finite
types are equivalent to cardinal finite types, so we will work with the latter
from now on.
Our first task is to show that cardinal finite types are closed under several
operations.
\subsection{Closure}
For the first three closure proofs, we only consider split enumerability:
as it is the strongest of the finiteness predicates, we can derive the other
closure proofs from it.

>>>>

%   We formalise our work in Cubical Agda, where we implement a library for proof
%   search (including combinators for level-polymorphic fully generic currying).
%   Through this library we demonstrate a number of uses for the computational
%   content of the univalence axiom, including searching for and synthesising
%   functions.
% \end{abstract}
\include{sections/introduction}
\include{sections/finiteness-predicates}
\include{sections/topos}
\include{sections/countable-predicates}
\include{sections/search}
\include{sections/related-work}

% \begin{acks}
%   This work has been supported by the Science Foundation Ireland under the
%   following grant: 13/RC/2D94 to Irish Software Research Centre.
% \end{acks}
